An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-10 EconGrowth.Rnw
Now forcing figure positions with "H". Began writing text for single-factor economic model analysis.

\section{The Facts}

Figure \ref{fig:Factors_Lattice_Graph} shows the ``facts'' as they are known.

<<Factors_Lattice_Graph, fig.width=figWidth, fig.height=figHeight, fig.cap="The facts. Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).">>=
createFactorsLatticeGraph <- function(){
  dataTable <- loadData("All")
  graphType <- "b" #b is for both line and symbol
  lineTypes <- c(0, 1, 5, 2, 4, 1) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(0, 2, 2, 1, 1, 1) #line widths. 0 means no line.
  colors <- c("black", "black", "black", "red", "blue", "darkorange") #line and symbol colors
  symbols <- c(1, NA, NA, NA, NA, NA)  #NA gives no symbol.
  graph <- xyplot(iGDP+iCapStk+iLabor+iQ+iX+iU ~ Year | Country, data=dataTable,
                  type = graphType,
                  index.cond = list(countryOrderForGraphs), #orders the panels.
                  layout = layoutSpec, 
                  strip = strip.custom(factor.levels=countryNamesAlph, bg="white"), #Sets strip parameters
                  as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                  lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                  pch = symbols, col.symbol = colors, #Controls symbol parameters
                  key=list(text=list(c("$y$", "$k$", "$l$", "$q$", "$x$", "$u$")),
                           type=graphType,
                           cex=keyTextSize, #controls size of text in the key
                           lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                           pch=symbols, col=colors, #controls symbol (plot characters) types
                           columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                  scales=list(cex=scaleTextSize, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              y=list(relation="free", #allows each axis to be different
                                     at=c(1,2,3,4,5,6,7,8,9,10) #tick mark locations
                                     )
                              ),
                  ylim=list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)), #y axis limits
                  xlab="", ylab="Indexed (1980=1 or 1991=1)") #axis labels
  return(graph)
}
createFactorsLatticeGraph()
@

Figure \ref{Factors_Lattice_Graph} shows that GDP ($y$), capital stock ($k$), labor ($l$), and the various energy types ($q$, $x$, and $u$) track each other quite closely. Covariances among these variables are quite high as shown in Tables \ref{tab:Covar_US} through \ref{tab:Covar_ZM} show.

@@

<<CD_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (without energy) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph()
@

@@

<<CDq_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("Q")
@

@@

<<CDx_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("X")
@

@@

<<CDu_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $u$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("U")
@

@@

<<CD_GDP_Lattice_Graph, fig.width=figWidth, fig.height=figHeight, fig.cap="Cobb-Douglas results.">>=
createCDLatticeGraph()
@

>>>>

\section{Historical Data}

Figure \ref{fig:Factors_Lattice_Graph} shows historical data. GDP ($y$), capital stock ($k$), labor ($l$), and the various energy types ($q$, $x$, and $u$) track each other closely. Covariances among these variables are rather high as shown in Tables \ref{tab:Covar_US} through \ref{tab:Covar_ZM}.

<<Factors_Lattice_Graph, fig.width=figWidth, fig.height=figHeight, fig.pos="H", fig.cap="The facts. Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).">>=
createFactorsLatticeGraph <- function(){
  dataTable <- loadData("All")
  graphType <- "b" #b is for both line and symbol
  lineTypes <- c(0, 1, 5, 2, 4, 1) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(0, 2, 2, 1, 1, 1) #line widths. 0 means no line.
  colors <- c("black", "black", "black", "red", "blue", "darkorange") #line and symbol colors
  symbols <- c(1, NA, NA, NA, NA, NA)  #NA gives no symbol.
  graph <- xyplot(iGDP+iCapStk+iLabor+iQ+iX+iU ~ Year | Country, data=dataTable,
                  type = graphType,
                  index.cond = list(countryOrderForGraphs), #orders the panels.
                  layout = layoutSpec, 
                  strip = strip.custom(factor.levels=countryNamesAlph, bg="white"), #Sets strip parameters
                  as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                  lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                  pch = symbols, col.symbol = colors, #Controls symbol parameters
                  key=list(text=list(c("$y$", "$k$", "$l$", "$q$", "$x$", "$u$")),
                           type=graphType,
                           cex=keyTextSize, #controls size of text in the key
                           lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                           pch=symbols, col=colors, #controls symbol (plot characters) types
                           columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                  scales=list(cex=scaleTextSize, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              y=list(relation="free", #allows each axis to be different
                                     at=c(1,2,3,4,5,6,7,8,9,10) #tick mark locations
                                     )
                              ),
                  ylim=list(c(1,10), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4), c(1,4)), #y axis limits
                  xlab="", ylab="Indexed (1980=1 or 1991=1)") #axis labels
  return(graph)
}
createFactorsLatticeGraph()
@

\section{Single-factor Models}

<<single-factor functions, eval=TRUE>>=

We begin by fitting single-factor economic models of the form

\begin{equation} \label{eq:Single_Factor_Generic}
  y = a^{\lambda (t-t_0)}f^{m}
\end{equation}

\noindent to historical data where the factor of production ($f$) is any of $k$, $l$, $q$, $x$, or $u$ and the exponent ($m$) is $\alpha$ for $k$, $\beta$ for $l$, and $\gamma$ for $q$, $x$, or $u$.

@@

<<CD_Params_Graph, fig.width=figWidth, fig.height=5, fig.pos="H", fig.cap="Cobb-Douglas (without energy) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph()
@

@@

<<CDq_Params_Graph, fig.width=figWidth, fig.height=5, fig.pos="H", fig.cap="Cobb-Douglas (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("Q")
@

@@

<<CDx_Params_Graph, fig.width=figWidth, fig.height=5, fig.pos="H", fig.cap="Cobb-Douglas (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("X")
@

@@

<<CDu_Params_Graph, fig.width=figWidth, fig.height=5, fig.pos="H", fig.cap="Cobb-Douglas (with $u$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("U")
@

@@

<<CD_GDP_Lattice_Graph, fig.width=figWidth, fig.height=figHeight, fig.pos="H", fig.cap="Cobb-Douglas results.">>=
createCDLatticeGraph()
@
